# Insert chapter 2 title here

*Describe the work you have done this week and summarize your learning.*

- Describe your work and results clearly. 
- Assume the reader has an introductory course level understanding of writing and reading R code as well as statistical methods.
- Assume the reader has no previous knowledge of your data or the more advanced methods you are using.

```{r}
date()
```

# 1. Dataset description

```{r}
library(tidyverse)
learning2014 <- read_csv('data/learning2014.csv')
str(learning2014)
```

"learning2014" dataset is a subset of "JYTOPKYS3" dataset containing 7 variables:

1. gender
2. age
3. attitude
4. deep (deep questions): combined of 12 variables from the original data set (starting with "D" letter)
5. stra (strategic questions): combined of 8 variables from the original data set (starting with "ST")
6. suf (surface questions): combined of 12 variables from the original data set (starting with "SU")
7. points: from statistics course exam


"deep", "stra", and "suf" are calculated by combining and averaging their corresponding variables.


# 2. Graphical overview


```{r}
# Access the gglot2 library
library(ggplot2)

summary(learning2014)

learning2014 %>% 
  ggplot(aes(x = gender)) +
  geom_bar()

learning2014 %>% 
  ggplot(aes(x = age)) +
  geom_histogram()

learning2014 %>% 
  ggplot(aes(x = attitude)) +
  geom_histogram()

learning2014 %>% 
  ggplot(aes(x = deep)) +
  geom_histogram()

learning2014 %>% 
  ggplot(aes(x = stra)) +
  geom_histogram()

learning2014 %>% 
  ggplot(aes(x = surf)) +
  geom_histogram()

learning2014 %>% 
  ggplot(aes(x = points)) +
  geom_histogram()

```

Gender graph obviously shows that the number of female participants are almost twice the number of males.
According to the age graph, majority of participants are 20-25 years old.
The attitude graph seems to follow a normal distribution which indicate most of participant have a moderate attitude toward statistics.
Comparing the mean of "deep" and the "surf", it shows the students tend to learn the course material more deeply, instead of surface studying.


# 3. Fitting regression line


```{r}
# Regression model with three explantory variables 
my_model_1 <- lm(points ~ attitude + stra + deep, data = learning2014)
summary(my_model_1)

# Excluding "deep" variable due to its insignificant relationship with  "points"
my_model_2 <- lm(points ~ attitude + stra, data = learning2014)
summary(my_model_2)

# Excluding "stra" variable as well
my_model_3 <- lm(points ~ attitude, data = learning2014)
summary(my_model_3)


```

In the first  model, we have used three explanatory variables of "attitude", "stra",
and "deep". According to the summary, there is no significant relationship between "deep"
and "points". Thus, we removed "deep" variable from the model, and fit the model again with
"attitude" and "stra" as the explanatory variables; still, it seems "stra" has not statistically
significant relationship with our target variable, "points". So, we fit the model once again
with only "attitude" as its explanatory variable. The summary shows a significant linear relationship
between "attitude" and "points" which means those who have positive attitude toward statistics, get
better points in this course.


# 4. Parameter interpretaion

Using the summary of fitted model, multiple R-squared in all three level, is low which indicates a weak relationship between our explanatory variables and points. However only one of these weak estimating relationship (between attitude and points) is significant.

# 5. diagnostic plots

```{r}
library(ggfortify)
autoplot(my_model_3)
```
The assumptions of linear regression are: 
- The relationship between the predictor (x) and the outcome (y) is assumed to be linear.
- The residual errors are assumed to be normally distributed.
- The residuals are assumed to have a constant variance (homoscedasticity)
- residuals error terms are independent.

We have checked the above assumptions by diagnostic plots. Residual vs Fitted plot is without pattern that shows the relationship between variables (attitude and points) is linear. Normal Q-Q plot follows a straight line which shows residuals have normally distributed. From Residuals vs Leverage plot, we see there is only three influential/extreme values in the data (#71, #35, #56) 

